Modeling the impacts of park access on health outcomes: a choice-based accessibility approach

### 2019-03-22 manuscript.Rnw
Minor edits for submittal

\author{Gregory S. Macfarlane\affilnum{1}, Nico Boyd\affilnum{2},
John E. Taylor\affilnum{2}, and Kari Watkins\affilnum{2}}
\runninghead{Boyd, et al.}

\affiliation{\affilnum{1}Civil and Environmental Engineering Department, Brigham Young University\\
\affilnum{2}School of Civil and Environmental Engineering, Georgia Institute of Technology}

\corrauth{Gregory Macfarlane\\
430 Engineering Building\\
Brigham Young University\\
Provo, UT 84602\\
USA}
\begin{abstract}
Recent research has underscored the potential for public green spaces to
influence individual and societal health outcomes, but
empirical measurements of such influences have yielded mixed results to date, with
particular disagreement surrounding how access to parks ought to be defined while
controlling alternate explanations. In this paper we present a comprehensive
measure of park accessibility drawn from random utility choice theory, which avoids
arbitrary assertions of proximity while incorporating potentially numerous
amenities and attributes of both the parks and the population. We apply this
choice-based accessibility measure to correlate Census tract-level obesity and
physical activity rate estimates from the Centers for Disease Control and
Prevention's 500 Cities project with tract-level American Community Survey
socioeconomic data in New York City, paired with geographic parks data from New
York City and geolocated tweet data collected from the Twitter API as a metric of
a park's activity levels. Controlling for the socioeconomic variables and
spatially correlated error terms, we show a positive and significant
relationship between park access and physical activity rates, and a clear and
significant negative relationship between park access and obesity rates. In
doing so, this research contributes a new, more comprehensive modeling approach
for measuring the impact of park access on health, and may improve our
understanding of the role  parks can serve in furthering public health
objectives.
\end{abstract}

@@

<<map, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggmap)
map <- ggmap(get_map(unname(st_bbox(tracts)), source = "stamen", maptype = "toner-lite"),
             extent = "device")
map +
  geom_sf(data = tracts, inherit.aes=FALSE,
          aes(fill = cut(access_ls, breaks = c(-Inf, -3, -2, -1, 0, 2, 3, Inf))),
          alpha = 0.8, color = NA) +
  scale_fill_brewer("Normalized Log-sum", palette = "GnBu")
@
  \caption{Normalized choice-based park accessibility log-sum values in New York City}
  \label{fig:map}
\end{figure*}

@@

<<tweets_map, echo=FALSE, message=FALSE, warning=FALSE>>=
library(ggmap)
map <- ggmap(get_map(unname(st_bbox(tracts)), source = "stamen", maptype = "toner-lite"),
             extent = "device")
map +
  geom_sf(data = tracts %>%
            mutate(diff = tweets_ls - access_ls,
                   pct_diff = diff / access_ls * 100),
          inherit.aes=FALSE,
          aes(fill = cut(pct_diff, quantile(pct_diff,
                                            probs = seq(0, 1, by = 0.1)))),
          alpha = 0.8, color = NA) +
  scale_fill_brewer("Percent Change", palette = "PiYG")
@
  \caption{Percent change in relative accessibility to parks when accounting for Twitter
  activity.}
  \label{fig:tweets_map}
\end{figure*}

>>>>

\author{Gregory S. Macfarlane\affilnum{1}, Nico Boyd\affilnum{2},
John E. Taylor\affilnum{2}, and Kari Watkins\affilnum{2}}
\runninghead{Modeling health outcomes via choice-based accessibility to parks}

\affiliation{\affilnum{1}Civil and Environmental Engineering Department, Brigham Young University\\
\affilnum{2}Fehr \& Peers, Los Angeles, California\\
\affilnum{3}School of Civil and Environmental Engineering, Georgia Institute of Technology}

\corrauth{Gregory Macfarlane: 430 Engineering Building, Provo, UT 84602 USA}
\email{gregmacfarlane@byu.edu}

\begin{abstract}
Recent research has underscored the potential for public green spaces to
influence individual and societal health outcomes, but
empirical measurements of such influences have yielded mixed results to date, with
particular disagreement surrounding how access to parks ought to be defined while
controlling alternate explanations. In this paper we present a comprehensive
measure of park accessibility drawn from random utility choice theory, which avoids
arbitrary assertions of proximity while incorporating potentially numerous
amenities and attributes of both the parks and the population. We apply this
choice-based accessibility measure to correlate Census tract-level obesity and
physical activity rate estimates from the Centers for Disease Control and
Prevention's 500 Cities project with tract-level American Community Survey
socioeconomic data in New York City, paired with geographic parks data from New
York City and geolocated tweet data collected from the Twitter API as a metric of
a park's activity levels. Controlling for the socioeconomic variables and
spatially correlated error terms, we show a positive and significant
relationship between park access and physical activity rates, and a clear and
significant negative relationship between park access and obesity rates. In
doing so, this research contributes a new, more comprehensive modeling approach
for measuring the impact of park access on health, and may improve our
understanding of the role  parks can serve in furthering public health
objectives.
\end{abstract}

\keywords{accessibility, spatial modelling, green space}

@@

<<map, echo=FALSE, message=FALSE, warning=FALSE, dev = 'tikz', fig.height=4>>=
library(ggmap)
map <- ggmap(get_map(unname(st_bbox(tracts)), source = "stamen", maptype = "toner-lite"),
             extent = "device")
map +
  geom_sf(data = tracts, inherit.aes=FALSE,
          aes(fill = cut(access_ls, breaks = c(-Inf, -3, -2, -1, 0, 2, 3, Inf))),
          alpha = 0.8, color = NA) +
  scale_fill_brewer("Normalized Log-sum", palette = "GnBu")
@
  \caption{Normalized choice-based park accessibility log-sum values in New York City.}
  \label{fig:map}
\end{figure*}

@@

<<tweets_map, echo=FALSE, message=FALSE, warning=FALSE, dev = 'tikz', fig.height=4>>=
library(ggmap)
map <- ggmap(get_map(unname(st_bbox(tracts)), source = "stamen", maptype = "toner-lite"),
             extent = "device")
map +
  geom_sf(data = tracts %>%
            mutate(diff = tweets_ls - access_ls,
                   pct_diff = diff / access_ls * 100),
          inherit.aes=FALSE,
          aes(fill = cut(pct_diff, quantile(pct_diff,
                                            probs = seq(0, 1, by = 0.1)))),
          alpha = 0.8, color = NA) +
  scale_fill_brewer("Percent Change", palette = "PiYG")
@
  \caption{Percent change in relative accessibility to parks when accounting for Twitter
  activity.}
  \label{fig:tweets_map}
\end{figure*}
